---
title: "Food and Transportation"
author: "Dan Price"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
geometry: margin=1in
fontsize: 11pt
documentclass: article
header-includes: 
  - \usepackage{tikz}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Preliminaries -- very broad
https://kbroman.org/knitr_knutshell/pages/Rmarkdown.html for options on the knit - results="hide"; echo=FALSE; include=FALSE, etc.

```{r prelims}
library(tidyr)
library(dplyr)
library(stringr)
library(data.table)
library(sf)
maindir = "~/Downloads/UH_OneDrive/OneDrive\ -\ University\ Of\ Houston/Social\ Network\ Hypergraphs/" #Dan at home
#maindir = "~/OneDrive\ -\ University\ Of\ Houston/Social\ Network\ Hypergraphs/" #Dan at work
housingdir = paste0(maindir,"HCAD/")
houstondatadir = paste0(maindir,"HoustonCityData/") 
censusdir = paste0(maindir,"Census/") 
vintage = "2022"
housingStockFromRDS = TRUE 
#numberOfCores = 1
state = 48 #48 Texas; 22 Louisiana
county = 201 #8 county region: 201 Harris; 157 Fort Bend; 167 Galveston; 039 Brazoria; 071 Chambers; 291 Liberty; 339 Montgomery; 473 Waller ; other place FIPS are longer
tract = "*"
censuskey <- readLines(paste0(censusdir, "2017", "/key"))
#HCAD has more census blocks than county=="201" - have to decide on best policy
geo_vintage <- "2010" #should be able to match vintage, depending on date.
source('BaseScripts/Census_Data.R') #this is just a quick thing you shouldn't need
```
#Notes on where I got files
from http://pdata.hcad.org/GIS/index.html
got Parcels 2015, Hwy, School, City, Sub_poly (subdivisions), TIRZ
on 12/27/2019 got 2017 of Parcels, and most recent of others

on 7/15/2022, got 2005 (no personal property data), 2010, 2020, 2021, 2022 downloads from https://hcad.org/pdata/pdata-property-downloads.html#

got 2015 acct information from: http://pdata.hcad.org/download/2015.html - 2005 - 2020 is available.
 PP-files, real_account, and real_building
downloaded 2017 from http://pdata.hcad.org/download/2017.html on Dec. 27, 2019.

http://hcad.org/hcad-resources/hcad-appraisal-codes/
http://hcad.org/hcad-resources/hcad-appraisal-codes/hcad-building-style-codes/ also exists, but has some diffs

code descriptions for real are from: http://pdata.hcad.org/Desc/2015/code_desc_real.txt #updated https://pdata.hcad.org/data/cama/2015/code_desc_real.txt
and personal: http://pdata.hcad.org/Desc/2015/code_desc_personal.txt #updated: #https://pdata.hcad.org/data/cama/2015/code_desc_personal.txt
also exists, but has some diffs, perhaps because of year?


they also provide column names from a screenshot of their Access db: http://pdata.hcad.org/DB/Pdata_Fieldnames.pdf
and http://pdata.hcad.org/Desc/Layout_and_Length.txt
census geometry files were taken on 12/27/2019 from: https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.2017.html

```{r get HCAD parcels and geometry files from census}
#start with loading files we need from HCAD
#most recent parcels doesn't have the _Oct
HCAD_parcels <- st_read(paste0(housingdir, vintage, "/Parcels/Parcels.shp"),stringsAsFactors = FALSE)
#for 2022, had 9 unexpected geometry messages
#or
#HCAD_parcels <- st_read(paste0(housingdir, vintage, "/Parcels_", vintage, "_Oct/Parcels.shp"),stringsAsFactors = FALSE)
#then rename
HCAD_parcels <- HCAD_parcels %>%
  rename(account=HCAD_NUM) %>%
  select(account,LocAddr,city,zip,geometry,Stacked)
#HCAD_parcels_dt <- as.data.table(HCAD_parcels)
#table(HCAD_parcels_dt[,Stacked]) #not sure what to do with them - they aren't particularly correlated with valid or duplicated
#HCAD_parcels_clean <- unique(HCAD_parcels_dt, by = ('account'))
HCAD_parcels$valid <- st_is_valid(HCAD_parcels$geometry)
HCAD_valid <- HCAD_parcels[which(HCAD_parcels$valid==TRUE),]
HCAD_invalid <- HCAD_parcels[which(HCAD_parcels$valid==FALSE),]
HCAD_geom <- st_as_sf(HCAD_valid)
HCAD_geom <- st_transform(HCAD_geom, crs = 3857)
#get centroids - about 1% of the parcels cross tracts, so need it as a point and not full geometry; 
#according to census rules, the apt buildings should be in a single tract, but that may not always be true.
HCAD_geom$centroid <- st_centroid(HCAD_geom$geometry)
```

```{r add geoid}
#do on blocks, then just make tract id from geoid for blocks
censusblocks <- st_read(paste0(censusdir, geo_vintage, "/geo_census/cb_", geo_vintage, "_", state, "_bg_500k/cb_", geo_vintage, "_", state, "_bg_500k.shp"))
censusblocks <- st_transform(censusblocks, crs = 3857)
censusblocks$geoid_centroid <- st_centroid(censusblocks$geometry)
CensusBlockforHCADParcels <- st_within(HCAD_geom$centroid, censusblocks)
#unlist into vector
CensusBlockforHCADParcelsunlisted <- rapply(CensusBlockforHCADParcels,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
CensusBlockforHCADParcelsunlisted <- unlist(CensusBlockforHCADParcelsunlisted)
# add census geoid information to each parcel
HCAD_geom$geoid=censusblocks$GEOID[CensusBlockforHCADParcelsunlisted] 
HCAD_geom$tract <- substr(HCAD_geom$geoid,6,11)
censusblocks <- as.data.table(censusblocks[which(censusblocks$COUNTYFP=="201"),])
censusblocks[,("geoid"):=GEOID]
```

Adding a little bit for Vivienne
```{r food pantries in census_blocks}
food_pantries <- read.csv(paste0(houstondatadir,"2022/harriscounty_foodpantries.csv"), header = FALSE)
food_pantries <- food_pantries %>%
  rename(address=V1,latitude=V2,longitude=V3)
food_pantries <- st_as_sf(food_pantries, coords = c("longitude", "latitude"), 
                 crs = 4326)
food_pantries <- st_transform(food_pantries, crs = 3857)
CensusBlockforFoodPantries <- st_within(food_pantries, censusblocks)
#unlist into vector
CensusBlockforFoodPantriesunlisted <- rapply(CensusBlockforFoodPantries,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
CensusBlockforFoodPantriesunlisted <- unlist(CensusBlockforFoodPantriesunlisted)
# add census geoid information to each parcel
food_pantries$geoid=censusblocks$GEOID[CensusBlockforFoodPantriesunlisted] 
food_pantries$tract <- substr(food_pantries$geoid,6,11)
#perhaps try https://rdrr.io/cran/nngeo/man/st_connect.html instead????
#or: https://gis.stackexchange.com/questions/356768/nngeo-nearest-neighbor-with-distance-dataframe
food_pantries_distance <- st_distance(censusblocks$geoid_centroid,food_pantries$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
censusblocks$food_pantry_distance <- as.matrix(apply(food_pantries_distance,1,min,na.rm=TRUE))
V_cleaned <- read.csv(paste0(houstondatadir,"2022/V_cleanedupdata.csv"), header = TRUE)
V_cleaned <- as.data.table(V_cleaned)
V_cleaned[,("TRACTCE"):=substr(GEO_ID,15,21)]
censusblocks <- censusblocks[which(censusblocks$COUNTYFP=="201"),]
food_pantries_geom <- merge(censusblocks,V_cleaned,by="TRACTCE")
food_pantries_geom <- st_transform(food_pantries_geom,crs=4326)
food_pantries <- st_transform(food_pantries,crs=4326)
#st_write(food_pantries_geom,"~/Downloads/food_pantries_geom.csv",driver = "CSV")
#st_write(food_pantries_geom,"~/Downloads/food_pantries_geom.geojson",driver="GeoJSON")
food_pantries$longitude <- st_coordinates(food_pantries)[,1]
food_pantries$latitude <- st_coordinates(food_pantries)[,2]
#st_write(food_pantries,"~/Downloads/food_pantries.csv",driver = "CSV")
#st_write(food_pantries,"~/Downloads/food_pantries.geojson",driver="GeoJSON")
```

Will add all the administrative and other boundaries in a later version - have to redo

```{r get HCAD raw files for real for processing}
#add Hcad_geom from HCAD_geo.R - break up parts later and add tests

HCAD_real_build <- read.csv2(paste0(housingdir, vintage, "/Real_building_land/building_other.txt"),
        stringsAsFactors = FALSE,
        sep = "\t", header = FALSE, quote="",colClasses = c("V1"="character"))
#there are no account numbers in common between res_build and real_build, but some codes are in both

HCAD_real_build$V1 <- str_remove_all(HCAD_real_build$V1, " ")
HCAD_real_build <- HCAD_real_build %>% 
  rename(account=V1,improv_typ_real=V4,impr_mdl_cd_real=V5,
        bldg_value_real=V9,qual_descript_real=V14,date_built_real=V15,
        yr_remodel_real=V16,bldg_sf_real=V27,nbhd_factor_real=V28,
        bldg_name=V32,units=V33,  
        lease_rate=V35,occupancy_rate=V36,total_income=V37) %>%
  filter(!row_number()==1) %>%
  select(account,improv_typ_real,impr_mdl_cd_real,bldg_value_real,qual_descript_real,
         date_built_real,yr_remodel_real,bldg_sf_real,nbhd_factor_real,
         bldg_name,units,lease_rate,occupancy_rate,total_income) %>%
  group_by(account) %>%
  mutate(dup_account_real = row_number() > 1) %>% 
  ungroup()
HCAD_real_build_dt <- as.data.table(HCAD_real_build)
HCAD_geom_dt <- as.data.table(HCAD_geom)
HCAD <- HCAD_real_build_dt[HCAD_geom_dt,on=.(account)]
nrow(HCAD[improv_typ_real=="4347"])==nrow(HCAD_real_build_dt[improv_typ_real=="4347"])
#need to add the distances from HCAD_geo
busstops <- readRDS(paste0(houstondatadir,vintage,"/Metro_stops/busstops_blocks.RDS"))
busstops <- st_transform(busstops,crs=st_crs(HCAD_geom))
#distance to busstops for each block group
block_bus_distance <- st_distance(censusblocks$geoid_centroid,busstops$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
bus_distance <- as.data.table(block_bus_distance)
censusblocks$bus_distance <- apply(bus_distance[,1:length(colnames(bus_distance))], 1, min)

#already have the within geoid marked by the fact that theses are in HCAD
fast_food_HCAD <- HCAD[improv_typ_real%in%c("4323","4325"),] #food stands (130) and fast food (1,627)
super_market_HCAD <- HCAD[improv_typ_real=="4347",]
convenience_stores_HCAD <- HCAD[improv_typ_real%in%c("4324","4348","4335"),] #and conv. attached to gas pump and truck stop
shop_center_HCAD <- HCAD[improv_typ_real%in%c("4341","4342","4343","4344","4345"),] #includes things like Target in 4345
food_sources_HCAD <- HCAD[HCAD$improv_typ_real %in% c("4325","4323","4324","4335","4347","4348"),] %>% 
  distinct(account, .keep_all = TRUE)

#for convenience stores
food_source_distance <- st_distance(censusblocks$geoid_centroid,food_sources_HCAD$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
censusblocks$food_source_distance <- as.matrix(apply(food_source_distance,1,min,na.rm=TRUE))

fast_food_distance <- st_distance(censusblocks$geoid_centroid,fast_food_HCAD$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
censusblocks$fast_food_distance <- as.matrix(apply(fast_food_distance,1,min,na.rm=TRUE))

super_market_distance <- st_distance(censusblocks$geoid_centroid,super_market_HCAD$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
censusblocks$super_market_distance <- as.matrix(apply(super_market_distance,1,min,na.rm=TRUE))

convenience_stores_distance <- st_distance(censusblocks$geoid_centroid,convenience_stores_HCAD$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
censusblocks$convenience_stores_distance <- as.matrix(apply(convenience_stores_distance,1,min,na.rm=TRUE))

shop_center_distance <- st_distance(censusblocks$geoid_centroid,shop_center_HCAD$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
censusblocks$shop_center_distance <- as.matrix(apply(shop_center_distance,1,min,na.rm=TRUE))

censusblocks <- as.data.table(censusblocks) #update its indices


#for each county, can have sex_age by blck group - and thus pop by block group, too
vintage <- "2020" #most recent I could get at the block level is for the redistricting numbers
block_sex_by_age_race_data_from_census_tx <- censusData_byGroupName(censusdir, vintage, state, censuskey, 
                       groupname = "P1",county_num = county,
                       block="block_group",api_type="dec/pl",path_suff="est.csv")

tot_pop <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_001N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

tot_pop_DT <- as.data.table(tot_pop)
tot_pop_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("pop_total_block"):=tot_pop_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

white_tot <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_003N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

white_tot_DT <- as.data.table(sex_by_age_race)
white_tot_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("white_total_block"):=white_tot_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

black_tot <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_004N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

black_tot_DT <- as.data.table(sex_by_age_race)
black_tot_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("black_total_block"):=black_tot_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

aiaa_tot <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_005N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

aiaa_tot_DT <- as.data.table(sex_by_age_race)
aiaa_tot_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("aiaa_total_block"):=aiaa_tot_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

asian_tot <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_006N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

asian_tot_DT <- as.data.table(sex_by_age_race)
asian_tot_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("asian_total_block"):=asian_tot_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

nhopi_tot <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_007N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

nhopi_tot_DT <- as.data.table(sex_by_age_race)
nhopi_tot_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("nhopi_total_block"):=nhopi_tot_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

other_race_tot <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_008N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

other_race_tot_DT <- as.data.table(sex_by_age_race)
other_race_tot_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("other_race_total_block"):=other_race_tot_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

two_races_tot <- block_sex_by_age_race_data_from_census_tx %>%
  filter(name=="P1_009N") %>%
  pivot_longer(4:ncol(block_sex_by_age_race_data_from_census_tx),names_to = "GEOID", values_to = "number_sams")

two_races_tot_DT <- as.data.table(sex_by_age_race)
two_races_tot_DT[,("geoid"):=str_remove_all(GEOID,"_")]
censusblocks[,("two_races_total_block"):=two_races_tot_DT[.SD,list(as.integer(number_sams)),on=.(geoid)]]

```

```{r make summaries for food insecurity and metro availability}
censusblocks <- censusblocks %>%
  group_by(TRACTCE) %>%
  mutate(pop_total_tract = as.integer(sum(pop_total_block)),
         tract_avg_bus = as.integer(sum(bus_distance*pop_total_block)/pop_total_tract),
         tract_avg_food = as.integer(sum(food_source_distance*pop_total_block)/pop_total_tract),
         tract_avg_fast_food = as.integer(sum(fast_food_distance*pop_total_block)/pop_total_tract),
         tract_avg_super_market = as.integer(sum(super_market_distance*pop_total_block)/pop_total_tract),
         tract_avg_convenience_stores = as.integer(sum(convenience_stores_distance*pop_total_block)/pop_total_tract),
         tract_avg_shop_center = as.integer(sum(shop_center_distance*pop_total_block)/pop_total_tract)
         )
censusblocks <- st_as_sf(censusblocks)
censusblocks <- st_transform(censusblocks,crs=4326)
saveRDS(censusblocks,"~/Downloads/censusblocks.RDS")
st_write(censusblocks,"~/Downloads/censusblocks.geojson",driver="GeoJSON")

```

Noah had things that didn't work for 2020 - they had been using the 2010 tract boundaries
Could try the crosswalk that the census provides, but since we have block level centroids, I'm just going to place them in the blocks that are there.

```{r crosswalk from 2010 to 2020 tract boundaries}
censusblocks <- readRDS("~/Downloads/censusblocks.RDS")
censusblocks <- st_transform(censusblocks,crs=3857)
censusblocks_19 <- st_read(paste0(censusdir, "2019/geo_census/cb_2019_", state, "_bg_500k/cb_2019_", state, "_bg_500k.shp"))
censusblocks_19 <- st_transform(censusblocks_19,crs=3857)
#censusblocks$geoid_centroid <- st_transform(censusblocks$geoid_centroid,crs=3857) #or censusblocks_19$geoid_centroids??
#find which TRACTCE each one is in, and then run the dyplyr from above with TRACTCE19 as the group
CensusBlockfor19 <- st_within(censusblocks$geoid_centroid, censusblocks_19)
#unlist into vector
CensusBlockfor19unlisted <- rapply(CensusBlockfor19,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
CensusBlockfor19unlisted <- unlist(CensusBlockfor19unlisted)
censusblocks$geoid_19=censusblocks_19$GEOID[CensusBlockfor19unlisted] 
censusblocks$tract_19 <- substr(censusblocks$geoid_19,6,11)
censusblocksnogeo <- censusblocks
censusblocksnogeo$geometry <- NULL
censusblocks_19 <- censusblocks_19[which(censusblocks_19$COUNTYFP=="201"),]
censusblocks_19$geoid_19 <- censusblocks_19$GEOID
censusblocks_19 <- left_join(censusblocksnogeo,censusblocks_19,by="geoid_19")
censusblocks_19$tract_19 <- substr(censusblocks_19$geoid_19,6,11)
censusblocks_19 <- censusblocks_19 %>%
  group_by(tract_19) %>%
  mutate(tract_pop_total = as.integer(sum(pop_total_block)),
         tract_avg_bus = as.integer(sum(bus_distance*pop_total_block)/pop_total_tract),
         tract_avg_food = as.integer(sum(food_source_distance*pop_total_block)/pop_total_tract),
         tract_avg_fast_food = as.integer(sum(fast_food_distance*pop_total_block)/pop_total_tract),
         tract_avg_super_market = as.integer(sum(super_market_distance*pop_total_block)/pop_total_tract),
         tract_avg_convenience_stores = as.integer(sum(convenience_stores_distance*pop_total_block)/pop_total_tract),
         tract_avg_shop_center = as.integer(sum(shop_center_distance*pop_total_block)/pop_total_tract)
         ) %>%
  distinct(tract_19, .keep_all = TRUE) %>%
  select(starts_with("tract_")) %>%
  select(!ends_with(".x") & !ends_with(".y")) %>%
  ungroup()
#now get tract geometry
geo_vintage <- "2019"
censustracts <- st_read(paste0(censusdir, geo_vintage, "/geo_census/cb_", geo_vintage, "_", state, "_tract_500k/cb_", geo_vintage, "_", state, "_tract_500k.shp"))
censustracts <- censustracts[which(censustracts$COUNTYFP=="201"),]
censustracts$tract_19 <- censustracts$TRACTCE
censusblocks_19$geometry <- NULL
censustracts_19 <- left_join(censustracts,censusblocks_19,by="tract_19")
#censusblocks_19 <- st_as_sf(censusblocks_19,sf_column_name = "geometry")
#censusblocks_19 <- st_transform(censusblocks_19,crs=4326)
saveRDS(censustracts_19,"~/Downloads/censustracts_19.RDS")
st_write(censustracts_19,"~/Downloads/censustracts_19.geojson",driver="GeoJSON")
```