---
title: "Making Sam Households"
author: "Dan Price"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
geometry: margin=1in
fontsize: 11pt
documentclass: article
header-includes: 
  - \usepackage{tikz}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Preliminaries -- very broad
```{r prelims}
source('~/Documents/Projects/SyntheticDataSet/BaseScripts/Census_Data.R')
library(tidyr)
library(dplyr)
library(stringr)
library(data.table)
#maindir = "~/University Of Houston/Price, Daniel M - Social Network Hypergraphs/"
maindir = "~/Downloads/UH_OneDrive/OneDrive\ -\ University\ Of\ Houston/Social\ Network\ Hypergraphs/" #Dan at home
#maindir = "~/Downloads/OneDrive\ -\ University\ Of\ Houston/Social\ Network\ Hypergraphs/" #Dan at work
housingdir = paste0(maindir,"HCAD/")
houstondatadir = paste0(maindir,"HoustonCityData/") 
censusdir = paste0(maindir,"Census/") 
vintage = "2019"
housingStockFromRDS = TRUE 
#numberOfCores = 1
state = 48 #48 Texas; 22 Louisiana
county = 201 #8 county region: 201 Harris; 157 Fort Bend; 167 Galveston; 039 Brazoria; 071 Chambers; 291 Liberty; 339 Montgomery; 473 Waller ; other place FIPS are longer
tract = "*"
Sam_seed = 135
#you don't need a censuskey if you're not pulling new files down; you can only use this one if you have correct access to the OneDrive
censuskey <- readLines(paste0(censusdir, "2017", "/key"))
```


For new file, and need a bit of a preamble

```{r basic household data download}
#block - sf1 - P15 and P16 (eth and race of hh) P18 (hhtype by race/eth) P22 (hhtype by age of hh) p28 (hhtype by hhsize by race/eth)
#plan - looks like P17 and P18 have much more detail than P28, so combine P17 with P28, then P18 with P28xP17
#then combine P16 with P28xP17xP18 and P15 with P28xP17xP18xP16   

#tract - sf2 - (Tenure by hhtype by age of hh - HCT9) - more age_groups and has tenure
```

```{r hh_relation download block}

dec_hh_relation_data_from_census_10 <- censusData_byGroupName(censusdir, vintage="2010", state, censuskey, 
                                                                groupname = "P29",county_num = "201",
                                                                block="block_group",api_type="dec/sf1",path_suff="est.csv")
```



The redistricting data collected in 2010 reports population estimates in yet a different format. It has one table that has races and ethnicity data for everyone over 18 and one that has it for everyone, with no ages attached. The race data gives more details on the composition of "Two or More Races" but nothing else new. The ethnicity file gives block level data for Hispanic or Latino who don't identify as White. Gives slightly more contour, but not much, and risks big differences on age and ethnicity or race per block group. (Although some contour is better than none!)

```{r pl data downloaded}
    dec_eth_block_data_from_census_10 <- censusData_byGroupName(censusdir, vintage="2010", state, censuskey, 
                                                                      groupname = "P2",county_num = "201",
                                                                      block="block_group",api_type="dec/pl",path_suff="est.csv")
    dec_race_block_data_from_census_10 <- censusData_byGroupName(censusdir, vintage="2010", state, censuskey, 
                                                                groupname = "P1",county_num = "201",
                                                                block="block_group",api_type="dec/pl",path_suff="est.csv")
    dec_eth_over18_block_data_from_census_10 <- censusData_byGroupName(censusdir, vintage="2010", state, censuskey, 
                                                                      groupname = "P4",county_num = "201",
                                                                      block="block_group",api_type="dec/pl",path_suff="est.csv")
    dec_race_over18_block_data_from_census_10 <- censusData_byGroupName(censusdir, vintage="2010", state, censuskey, 
                                                                groupname = "P3",county_num = "201",
                                                                block="block_group",api_type="dec/pl",path_suff="est.csv")
 
    
```




Then have the tract level details join into the bg representation? Should discuss whether to start with the household stuff - check if it has the PCT level of detail on ethnicity anywhere... 
If not, can just save the demographic_bg to the OneDrive and start from there.




https://www2.census.gov/programs-surveys/decennial/2010/technical-documentation/complete-tech-docs/summary-file/sf1.pdf
put in households
add P15 or P16, then other hh stuff that can be added without problems, then add people in
also look at H9
P21 should also be good; p29, p30...

have to think about how family info is different from hh

group quarters are needed b/c hh don't include them
```{r group quarters to add}
dec_group_quarters_block_data_from_census_10 <- 
  censusData_byGroupName(censusdir, vintage="2010", state, censuskey, 
                         groupname = "P5",county_num = "201",
                         block="block_group",api_type="dec/pl",path_suff="est.csv")
```



expand from census

rearrange so that they all add to same whole and then figure out how to distribute them among that space

tests of each step

discussion of power sets vs. categories? 

redoing HCAD first to help with generation of 2011 - a space to move into...


