---
title: "Structuring HCAD for Sam"
author: "Dan Price"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
geometry: margin=1in
fontsize: 11pt
documentclass: article
header-includes: 
  - \usepackage{tikz}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Preliminaries -- very broad
https://kbroman.org/knitr_knutshell/pages/Rmarkdown.html for options on the knit - results="hide"; echo=FALSE; include=FALSE, etc.

```{r prelims}
library(tidyr)
library(dplyr)
library(stringr)
library(data.table)
library(sf)
library(lwgeom)
#maindir = "~/University Of Houston/Price, Daniel M - Social Network Hypergraphs/"
maindir = "~/Downloads/UH_OneDrive/OneDrive\ -\ University\ Of\ Houston/Social\ Network\ Hypergraphs/" #Dan at home
#maindir = "~/Downloads/OneDrive\ -\ University\ Of\ Houston/Social\ Network\ Hypergraphs/" #Dan at work
housingdir = paste0(maindir,"HCAD/")
houstondatadir = paste0(maindir,"HoustonCityData/") 
censusdir = paste0(maindir,"Census/") 
vintage = "2022"
housingStockFromRDS = TRUE 
#numberOfCores = 1
state = 48 #48 Texas; 22 Louisiana
county = 201 #8 county region: 201 Harris; 157 Fort Bend; 167 Galveston; 039 Brazoria; 071 Chambers; 291 Liberty; 339 Montgomery; 473 Waller ; other place FIPS are longer
#HCAD has more census blocks than county=="201" - have to decide on best policy
geo_vintage <- "2021" #most recent - it should be able to match vintage, depending on date.
```
#Notes on where I got files
https://public.hcad.org/resources/statecode.asp

from http://pdata.hcad.org/GIS/index.html - no longer maintained there.
got Parcels 2015, Hwy, School, City, Sub_poly (subdivisions), TIRZ
on 12/27/2019 got 2017 of Parcels, and most recent of others

on 7/15/2022, got 2005 (no personal property data), 2010, 2020, 2021, 2022 downloads from https://hcad.org/pdata/pdata-property-downloads.html#

got 2015 acct information from: http://pdata.hcad.org/download/2015.html - 2005 - 2020 is available.
 PP-files, real_account, and real_building
downloaded 2017 from http://pdata.hcad.org/download/2017.html on Dec. 27, 2019.

http://hcad.org/hcad-resources/hcad-appraisal-codes/
http://hcad.org/hcad-resources/hcad-appraisal-codes/hcad-building-style-codes/ also exists, but has some diffs

code descriptions for real are from: http://pdata.hcad.org/Desc/2015/code_desc_real.txt #updated https://pdata.hcad.org/data/cama/2015/code_desc_real.txt
and personal: http://pdata.hcad.org/Desc/2015/code_desc_personal.txt #updated: #https://pdata.hcad.org/data/cama/2015/code_desc_personal.txt
also exists, but has some diffs, perhaps because of year?


they also provide column names from a screenshot of their Access db: http://pdata.hcad.org/DB/Pdata_Fieldnames.pdf (they took that down!)
and http://pdata.hcad.org/Desc/Layout_and_Length.txt (and took this down!)
census geometry files were taken on 12/27/2019 from: https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.2017.html

```{r get HCAD parcels and geometry files from census}
#censustracts <- st_read(paste0(censusdir, geo_vintage, "/geo_census/cb_", 
#                               geo_vintage, "_", state, "_tract_500k/cb_", geo_vintage, "_", state, "_tract_500k.shp"))
#censustracts <- st_transform(censustracts,crs = 3857) 
#start with loading files we need from HCAD
#most recent parcels doesn't have the _Oct
ifelse(file.exists(paste0(housingdir, vintage, "/Parcels/Parcels.shp")),
       HCAD_parcels <- st_read(paste0(housingdir, vintage, "/Parcels/Parcels.shp"),stringsAsFactors = FALSE,quiet = TRUE),
       HCAD_parcels <- st_read(paste0(housingdir, vintage, "/Parcels_", vintage, "_Oct/Parcels.shp"),stringsAsFactors = FALSE,quiet = TRUE)) #verbose = FALSE throws a weird error!! Just have to live with the printout, for now
#small number of warnings about the geometry - says they're saved as a collection, and it's hard to see if they end up being lost to the validity check later. There's a variable called "Stacked," but also seems uncorrelated to the warnings.
#then rename
HCAD_parcels <- HCAD_parcels %>%
  rename(account=HCAD_NUM) %>%
  select(account,LocAddr,city,zip,geometry)
#check for dups - 
nrow(HCAD_parcels)-length(unique(HCAD_parcels$account))
HCAD_parcels <- HCAD_parcels %>%
  group_by(account) %>%
  mutate(dup_rows = row_number()) %>%
  ungroup()
#on 2022, there were 7 duplicate rows, and they had matching geometries
#HCAD_parcels_dups[which(HCAD_parcels_dups$account=="1452480010003"),]$geometry == HCAD_parcels_dups[which(HCAD_parcels_dups$account=="1452480010003"),]$geometry
HCAD_parcels <- HCAD_parcels[which(HCAD_parcels$dup_rows==1),]
HCAD_parcels$dup_rows <- NULL
HCAD_parcels$valid <- st_is_valid(HCAD_parcels$geometry)
HCAD_parcels[which(HCAD_parcels$valid==FALSE),]$geometry <-
  st_make_valid(HCAD_parcels[which(HCAD_parcels$valid==FALSE),]$geometry)
HCAD_geom <- st_as_sf(HCAD_parcels)
HCAD_geom <- st_transform(HCAD_geom, crs = 3857)
#get centroids - about 1% of the parcels cross tracts, so need it as a point and not full geometry; 
#according to census rules, the apt buildings should be in a single tract, but that may not always be true.
HCAD_geom$centroid <- st_centroid(HCAD_geom$geometry)
```

Add the appropriate geoid (block and tract numbers included):
```{r add geoid}
#do on blocks, then just make tract id from geoid for blocks
censusblocks <- st_read(paste0(censusdir, geo_vintage, "/geo_census/cb_", geo_vintage, "_", state, "_bg_500k/cb_", geo_vintage, "_", state, "_bg_500k.shp"))
censusblocks <- st_transform(censusblocks,st_crs(HCAD_geom))
censusblocks$geoid_centroid <- st_centroid(censusblocks$geometry)
CensusBlockforHCADParcels <- st_within(HCAD_geom$centroid, censusblocks)
#unlist into vector
CensusBlockforHCADParcelsunlisted <- rapply(CensusBlockforHCADParcels,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
CensusBlockforHCADParcelsunlisted <- unlist(CensusBlockforHCADParcelsunlisted)
# add census geoid information to each parcel
HCAD_geom$geoid=censusblocks$GEOID[CensusBlockforHCADParcelsunlisted] 
HCAD_geom$tract <- substr(HCAD_geom$geoid,6,11)
censusblocks <- as.data.table(censusblocks)
paste0("There are ",length(unique(HCAD_geom$geoid)) -
                             length(unique(censusblocks[COUNTYFP=="201",GEOID])),
       " more block groups in HCAD than in the census definition of Harris County.")
paste0("There are ",length(unique(HCAD_geom$tract)),
       " tracts in HCAD.")
#check them on tracts, too
#write out missing to geojson and look at kepler.gl - for now, will just add the extra blocks to be complete
censusblocks <- censusblocks[GEOID%in%unique(HCAD_geom$geoid)]
censusblocks[,("geoid"):=GEOID]
paste0("There are ",length(HCAD_geom[which(is.na(HCAD_geom$geoid)),])," NAs for geoid in the HCAD_geom file")
```
We add the 2010-19 geoid designations, too, since many of the matching data sets use it
```{r add geoid_19 - assuming that even the older ones might need it}
#do on blocks, then just make tract id from geoid for blocks
geo_vintage <- "2019"
censusblocks_19 <- st_read(paste0(censusdir, geo_vintage, "/geo_census/cb_", geo_vintage, "_", state, "_bg_500k/cb_", geo_vintage, "_", state, "_bg_500k.shp"))
censusblocks_19 <- st_transform(censusblocks_19,st_crs(HCAD_geom))
censusblocks_19$geoid_centroid <- st_centroid(censusblocks_19$geometry)
CensusBlockforHCADParcels <- st_within(HCAD_geom$centroid, censusblocks_19)
#unlist into vector
CensusBlockforHCADParcelsunlisted <- rapply(CensusBlockforHCADParcels,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
CensusBlockforHCADParcelsunlisted <- unlist(CensusBlockforHCADParcelsunlisted)
# add census geoid information to each parcel
HCAD_geom$geoid_19=censusblocks_19$GEOID[CensusBlockforHCADParcelsunlisted] 
HCAD_geom$tract_19 <- substr(HCAD_geom$geoid_19,6,11)
censusblocks_19 <- as.data.table(censusblocks_19)
paste0("There are ",length(unique(HCAD_geom$geoid_19)) -
                             length(unique(censusblocks_19[COUNTYFP=="201",GEOID])),
       " more block groups in HCAD than in the census definition of Harris County.")
paste0("There are ",length(unique(HCAD_geom$tract_19)),
       " more tracts in HCAD for the 2010-2019 geometries.")
#save only the censusblocks that relate to HCAD
censusblocks_19 <- censusblocks_19[GEOID%in%unique(HCAD_geom$geoid_19)]
censusblocks_19[,("geoid_19"):=GEOID]
paste0("There are ",length(HCAD_geom[which(is.na(HCAD_geom$geoid_19)),])," NAs for geoid_19 in the HCAD_geom file")
```

```{r add administrative boundaries}
#superneighborhoods - some changes in houstondata can happen - have to look at each dataset, since they don't consistently update by year
#it's only for city of Houston, so lots of Harris County will get NAs
superneighborhoods <- st_read(paste0(houstondatadir, "2022/HOUSTON_LIMITS_BOUNDARIES_PACKAGE/HOUSTON_LIMITS_BOUNDARIES_PACKAGE.shp"))
#2017 had different title, but same geometry
#superneighborhoods <- st_read(paste0(houstondatadir, "2017/COH_SUPER_NEIGHBORHOODS/COH_SUPER_NEIGHBORHOODS.shp"))
superneighborhoods <- st_transform(superneighborhoods, st_crs(HCAD_geom)) #HCAD is renamed from sf_HCAD in this run - can change
super_within <- st_within(HCAD_geom$centroid, superneighborhoods)
super_within_unlist <- rapply(super_within,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
super_within_unlist <- unlist(super_within_unlist)
HCAD_geom$superneighborhood=superneighborhoods$SNBNAME[super_within_unlist]
length(unique(HCAD_geom$superneighborhood))==length(unique(superneighborhoods$SNBNAME))+1 #b/c NA in HCAD

#city council for 2022 - some of the geometry is invalid and couldn't be fixed
#city_council <- st_read(paste0(houstondatadir, #"2022/City_Council/COH_ADMINISTRATIVE_BOUNDARY_-_MIL/COH_ADMINISTRATIVE_BOUNDARY_-_MIL.shp"))
#city_council <- st_transform(city_council, st_crs(HCAD_geom))
#city_council$valid <- st_is_valid(city_council$geometry) #- some weren't valid
#city_council[which(city_council$valid==FALSE),]$geometry <-
#  st_make_valid(city_council[which(city_council$valid==FALSE),]$geometry)
#council_within <- st_within(HCAD_geom$centroid,city_council)
#council_within_unlist <- rapply(council_within,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
#council_within_unlist <- unlist(council_within_unlist)
#HCAD_geom$council_district=city_council$DISTRICT[council_within_unlist]

#HISD 
hisd <- st_read(paste0(houstondatadir, "2022/HISD/Feeder_Pattern_2021_2022.shp"))
hisd <- st_transform(hisd, st_crs(HCAD_geom))
hisd_within <- st_within(HCAD_geom, hisd)
hisd_within_unlist <- rapply(hisd_within,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
hisd_within_unlist <- unlist(hisd_within_unlist)
HCAD_geom$HISD_High_School=hisd$High_Schoo[hisd_within_unlist]
HCAD_geom$HISD_Middle_School=hisd$Middle_Sch[hisd_within_unlist]
HCAD_geom$HISD_Elementary_School=hisd$Elementary[hisd_within_unlist]
length(unique(HCAD_geom$HISD_High_School)) == length(unique(hisd$High_Schoo)) + 1
length(unique(HCAD_geom$HISD_Elementary_School)) == length(unique(hisd$Elementary)) + 1
geo_vintage <- "2021"
other_isd <- st_read(paste0(censusdir, geo_vintage, "/geo_census/cb_", geo_vintage, "_", state, "_unsd_500k/cb_", geo_vintage, "_", state, "_unsd_500k.shp"))
other_isd <- st_transform(other_isd, st_crs(HCAD_geom))
other_isd_within <- st_within(HCAD_geom, other_isd)
other_isd_within_unlist <- rapply(other_isd_within,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
other_isd_within_unlist <- unlist(other_isd_within_unlist)
HCAD_geom$ISD=other_isd$NAME[other_isd_within_unlist]
paste0("There are ",length(unique(HCAD_geom$ISD))," ISDs in HCAD.")
paste0("The merge has left ",nrow(HCAD_geom[which(is.na(HCAD_geom$ISD)),])," parcels in HCAD without an ISD,which is similar to the number that didn't have LocAddr or city.")
#zipcodes are pulled from the HCAD site_addr_3 in the real_acct file
HCAD_geom_dt <- as.data.table(HCAD_geom)
```

```{r get HCAD raw files for processing}
#https://hcad.org/assets/uploads/pdf/pdataCodebook.pdf also in 2022
#state_codes are in: https://www.taxnetusa.com/research/texas/sptb.php (HCAD's own is incomplete: https://hcad.org/search/state-category-codes/)
#HCAD Site Codes (improv_typ) are in: https://hcad.org/hcad-resources/hcad-appraisal-codes/hcad-site-codes/

HCAD_res_build <- read.csv2(paste0(housingdir, vintage, "/Real_building_land/building_res.txt"),
        stringsAsFactors = FALSE,
        sep = "\t", header = TRUE, quote="",colClasses = c("acct"="character"))

#there are duplicate accounts (about 29k in 2022); just added total_sf in; seems to work from account side
HCAD_res_build$acct <- str_remove_all(HCAD_res_build$acct, " ")
HCAD_res_build <- HCAD_res_build %>% 
  rename(account=acct,state_code=property_use_cd,improv_typ=impr_tp,impr_mdl_cd=impr_mdl_cd,
         qual_descript=dscr,living_sf=im_sq_ft) %>%
  select(account,state_code,improv_typ,impr_mdl_cd,structure_dscr,
          qual_descript,date_erected,yr_remodel,
         living_sf,size_index) %>%
  group_by(account) %>%
  mutate(num_buildings = n(),
         total_living_sf = sum(as.numeric(living_sf),na.rm = TRUE)) %>%
  distinct(account, .keep_all = TRUE) %>%
  ungroup()

HCAD_res_build_dt <- as.data.table(HCAD_res_build)
#HCAD_res_geom_dt <- HCAD_geom_dt[HCAD_res_build_dt,on=.(account)]
#HCAD_res_geom <- HCAD_res_build_dt[HCAD_geom_dt,on=.(account)] 
#test <- table(HCAD_res_geom_dt$improv_typ)==
#  table(HCAD_res_build$improv_typ)
#length(test[test==FALSE])==0
#length(unique(HCAD_res_build_dt$account))==length(unique(HCAD_res_geom_dt$account))
#nrow(HCAD_res_build_dt)==nrow(HCAD_res_geom_dt)

HCAD_real_build <- read.csv2(paste0(housingdir, vintage, "/Real_building_land/building_other.txt"),
        stringsAsFactors = FALSE,
        sep = "\t", header = TRUE, quote="",colClasses = c("acct"="character"))

#there are no account numbers in common between res_build and real_build, but some codes are in both
#total_income may include non-rental income?
HCAD_real_build$acct <- str_remove_all(HCAD_real_build$acct, " ")
HCAD_real_build <- HCAD_real_build %>% 
  rename(account=acct,state_code=property_use_cd,improv_typ=impr_tp,
        qual_descript=dscr,bldg_sf=im_sq_ft,
        bldg_name=prop_nm,
        lease_rate=lease_rt,occupancy_rate=occ_rt,total_income=tot_inc) %>%
  select(account,state_code,improv_typ,impr_mdl_cd,qual_descript,
         date_erected,yr_remodel,bldg_sf,
         bldg_name,units,lease_rate,occupancy_rate,total_income) %>%
  group_by(account) %>%
  mutate(num_buildings = n(),
         total_bldg_sf = sum(as.numeric(bldg_sf),na.rm = TRUE)) %>% #units seems to already be summarized!!
  distinct(account, .keep_all = TRUE) %>%
  ungroup()
#correctional <- HCAD_real_build %>% filter(improv_typ=="4670") 
#30 correctional facilities, without number of rooms
HCAD_real_build_dt <- as.data.table(HCAD_real_build)
#HCAD_real_geom_dt <- HCAD_geom_dt[HCAD_real_build_dt,on=.(account)]
#test <- table(HCAD_real_geom_dt$improv_typ)==
#  table(HCAD_real_build_dt$improv_typ)
#length(test[test==FALSE])==0
#length(unique(HCAD_real_build_dt$account))==length(unique(HCAD_real_geom_dt$account))
#nrow(HCAD_real_build_dt)==nrow(HCAD_real_geom_dt)

HCAD_real_acct <- read.csv2(paste0(housingdir, vintage, "/Real_acct_owner/real_acct.txt"),
        stringsAsFactors = FALSE,
        sep = "\t", header = TRUE, quote="",colClasses = c("acct"="character"))
HCAD_real_acct$acct <- str_remove_all(HCAD_real_acct$acct, " ")
HCAD_real_acct <- HCAD_real_acct %>% 
  rename(account=acct,site_address_street=site_addr_1,
         site_address_city=site_addr_2,zip_code=site_addr_3,
         state_code=state_class,tot_bldg_sf=bld_ar,
         tot_land_sf=land_ar,land_value=land_val,
         improv_value=bld_val,xtras_value=x_features_val,market_value=tot_mkt_val) %>%
  select(account,site_address_street,site_address_city,zip_code,
         state_code,tot_bldg_sf,tot_land_sf,land_value,improv_value,
         xtras_value,market_value)
nrow(HCAD_real_acct)==length(unique(HCAD_real_acct$account))
##will use market_value to guestimate cost of apt.; here, other codes will also help decide (like LHI)
HCAD_real_acct_dt <- as.data.table(HCAD_real_acct)
#not_in_parcels <- HCAD_real_acct_dt[!account%in%unique(HCAD_geom_dt$account)]
#some of the above we're also not in parcels
#table(not_in_parcels$state_code)
#paste0("Of the ",nrow(not_in_parcels)," accounts without matching parcel geometries, ",
#       nrow(not_in_parcels[state_code=="X1"]), " are in government exempt. It seems to be things owned by one person and sitting on another person's lot.")
#in 2022, there are 66351 real accounts not in Parcels; not sure how to do them all.
#the mobile homes show up in the HCAD_res_geom, but without geometry - matching on site_addr is incomplete at best
#HCAD_res_acct_geom_dt <- HCAD_real_acct_dt[HCAD_res_geom_dt,on=.(account,state_code)]
#HCAD_real_acct_geom_dt <- HCAD_real_acct_dt[HCAD_real_geom_dt,on=.(account,state_code)]
#tests
#test <- table(HCAD_res_acct_geom_dt$improv_typ) == table(HCAD_res_geom_dt$improv_typ)
#length(test[test==FALSE])==0
#test <- table(HCAD_real_acct_geom_dt$improv_typ) == table(HCAD_real_geom_dt$improv_typ)
#length(test[test==FALSE])==0

#HCAD_owner_date <- read.csv2(paste0(housingdir, vintage, "/Real_acct_owner/deeds.txt"),
#                             stringsAsFactors = FALSE,
#                             sep = "\t", header = TRUE, quote="",colClasses = c("acct"="character"))
#want to get this eventually, for how recently people moved, but something weird about it, so skipping now
#remember to replace vintage with date, below
#HCAD_owner_date <- HCAD_owner_date %>%
#  rename(account=V1,purchase_date=V2) %>%
#  #distinct(account,.keep_all = TRUE) %>%
#  filter(!row_number()==1) %>%
#  select(account,purchase_date)  %>%
#  group_by(account) %>%
#  mutate(dup_account_acct = row_number() > 1) %>% 
#  ungroup()
##  mutate(
##    yrs_in_house = 2022 - as.integer(substr(as.character.Date(HCAD_owner_date$V2),7,10))
##  )
#HCAD <- full_join(HCAD,HCAD_owner_date,by="account")
#nrow_HCAD4 <- nrow(HCAD)

HCAD_exempt <- read.csv2(paste0(housingdir, vintage, "/Real_jur_exempt/jur_exempt.txt"),
                         stringsAsFactors = FALSE,
                         sep = "\t", header = TRUE, quote="",colClasses = c("acct"="character"))
#HCAD_exempt$V1 <- str_remove_all(HCAD_exempt$V1, " ") 
#exemptions give us some clues about owners, etc. that match up with ACS info
#cf. jur_exemption_dscr.txt in Real_jur_exempt
#THIS DOESN'T WORK; HAVE TO GO THROUGH THE LIST OF POSSIBLE EXEMPTS
HCAD_exempt_wide <- HCAD_exempt %>% 
  rename(account=acct) %>%
  group_by(account) %>%
  pivot_wider(id_cols = "account",names_from = "exempt_cat",values_from = "exempt_val",
              names_expand = TRUE,values_fn = list) %>%
  mutate(
    Total_val = ifelse(is.character(TOT[[1]][[1]][1]),TOT[[1]][[1]][1],"No Value"),
    
  )

#might be a better way to do it - most of them are null across all. 
  #    homestead = case_when(  #trying to get at who owns...
# #    #this has to somehow get at if any is true, then it triggers True - is there an "at least one TRUE"..??
  #      exempt_cat %in% c('RES','PAR') ~ TRUE
  #    ),
  #    vet = case_when(
  #      exempt_cat %in% c('V11','V12','V13','V14','V21','V22','V23','V24') ~ TRUE
  #    ),
  #    low_income_housing = case_when(
  #      exempt_cat == "LIH" ~ TRUE
  #    ),
  #    over_65 = case_when(
  #      exempt_cat %in% c('APO','OVR','POV','SUR') ~ TRUE
  #    ),
  #    surv_spouse = case_when(
  #      exempt_cat %in% c('SSD','STT','STX','SUR','VS1','VS2','VS3','VS4','VTX') ~ TRUE
  #    ),
  #    disabled = case_when(
  #      exempt_cat %in% c('PDS','DIS','APD','STX','V11','V12','V13','V14','V21','V22','V23','V24','VCH','VTX') ~ TRUE
  #    )
  #  ) %>%
  #distinct(account, .keep_all = TRUE) %>%
  #ungroup() 

HCAD_exempt_dt <- as.data.table(HCAD_exempt_wide)

#HCAD Site Codes (improv_typ) are in: https://hcad.org/hcad-resources/hcad-appraisal-codes/hcad-site-codes/
HCAD_fixtures <- read.csv2(paste0(housingdir, vintage, "/Real_building_land/fixtures.txt"),
                           stringsAsFactors = FALSE,
                           sep = "\t", header = TRUE, quote="",colClasses = c("acct"="character"))
#add counts for rooms and apartments
HCAD_fixtures <- HCAD_fixtures %>%
  rename(account=acct,bldg_num=bld_num,bldg_data=type,
         bldg_data_descr=type_dscr,num_units=units) %>%
  select(account,bldg_num,bldg_data,bldg_data_descr,num_units) %>%
  mutate(
    bedrooms = case_when(
      bldg_data == 'RMB ' ~ as.integer(num_units),
      bldg_data == 'AP0 ' ~ as.integer(0),
      bldg_data == 'AP1 ' ~ as.integer(1),
      bldg_data == 'AP2 ' ~ as.integer(2),
      bldg_data == 'AP3 ' ~ as.integer(3),
      bldg_data == 'AP4 ' ~ as.integer(4)
    ),
    total_rooms = case_when(
      bldg_data == 'RMT ' ~ as.integer(num_units)
    ),
    effcncy_apts = case_when(
      bldg_data == 'AP0 ' ~ as.integer(num_units)
    ),
    oneBR_apts = case_when(
      bldg_data == 'AP1 ' ~ as.integer(num_units)
    ),
    twoBR_apts = case_when(
      bldg_data == 'AP2 ' ~ as.integer(num_units)
    ),
    threeBR_apts = case_when(
      bldg_data == 'AP3 ' ~ as.integer(num_units)
    ),
    fourBR_apts = case_when(
      bldg_data == 'AP4 ' ~ as.integer(num_units),
      TRUE ~ as.integer(0)
    ),
    num_apts = case_when(  #maybe not use - seems to have a lot of missing!
      grepl('AP', bldg_data) ~ as.integer(num_units),
      #bldg_data == 'UNT ' ~ as.integer(num_units), #need to confirm it adds correctly
      TRUE ~ as.integer(0)
    )
  ) %>%
  filter(bedrooms>=0 | total_rooms>0 | effcncy_apts>0 | oneBR_apts>0 | 
           twoBR_apts>0 | threeBR_apts>0 | fourBR_apts>0 | num_apts>0 ) %>%
  group_by(account) %>%
  mutate(total_num_units = sum(num_apts)) %>%
  distinct(account,.keep_all = TRUE) %>%
  ungroup()
HCAD_fixtures_dt <- as.data.table(HCAD_fixtures)
#check to see how many are in res and how many in real
#test_real <- unique(HCAD_fixtures$account) %in% unique(HCAD_real_geom_dt$account)
#length(test_real[test_real==TRUE])
#HCAD_real_fixt_dt <- HCAD_real_build_dt[account %in% unique(HCAD_fixtures$account),]
```

```{r now for the joins}
HCADra_ex <- HCAD_real_acct_dt[HCAD_exempt_dt,on=.(account)]
HCADra_ex_miss <- HCAD_real_acct_dt[!account%in%unique(HCAD_exempt_dt$account),]
nrow(HCADra_ex[is.na(improv_value)])
HCADra_ex_geom <- HCADra_ex[HCAD_geom_dt,on=.(account)]
HCADra_ex_geom_miss <- HCADra_ex[!account%in%unique(HCAD_geom_dt$account),]
nrow(HCADra_ex_geom[is.na(improv_value)])
HCAD_res_fixt_dt <- HCAD_res_build_dt[account %in% unique(HCAD_fixtures$account),] #had a handful (21) not match!!
HCADra_ex_geom_fxt <- HCAD_res_fixt_dt[HCADra_ex_geom,on=.(account)] 
HCAD_real_fixt_dt <- HCAD_real_build_dt[account %in% unique(HCAD_fixtures$account),]
HCAD <- HCAD_real_fixt_dt[HCADra_ex_geom_fxt,on=.(account)]
#NEED to fix the .i on improv_typ and data_erected - I think it comes from doing two for res_build AND real_build
#MAYBE add them to each other first, then do join???
HCAD[,("num_residences"):=if_else(is.na(units),as.integer(1),as.integer(units))] #304 NAs, still



#HCAD_geom_dt <- HCAD_real_fixt_dt[HCAD_geom_dt,on=.(account)]
#paste0("Missing ",
#nrow(HCAD_real_fixt_dt[!is.na(improv_typ)])-nrow(HCAD_geom_dt[!is.na(improv_typ)]),
#"of the geometries for the fixtures for the real buildings - some of the mobile homes, etc.")
#not_in_parcels_real <- 
#then add to res file

#270k without matches!!!
nrow(HCAD[is.na(i.improv_typ) & is.na(improv_typ)])

#HCAD_fix_b4 <- nrow(HCAD_fixtures)
#HCAD_fixtures_dt <- as.data.table(HCAD_fixtures)
#HCAD5 <- HCAD_fixtures_dt[HCAD4,on=.(account)]
##there are lots of duplicate accounts - 
##expand so each apt has its own record
#HCAD6 <- data.table(HCAD5[rep(seq_len(dim(HCAD_fixtures_dt)[1])), 
#           as.integer(HCAD5$num_apts)+1, 
#            #as.integer(HCAD_fixtures$num_apts*.89)), #in case want to do occupancy this way
#            1:dim(HCAD5)[2], drop = FALSE], row.names = NULL)
##.89 is the overall Harris County occupancy - I have it in another place, too
#HCAD_fix_after <- nrow(HCAD6)

#add apt_ids
##HCAD_fixtures_single <- HCAD_fixtures %>%
##  group_by(account) %>%
##  mutate(
##    apt_id = paste0(account,rep(1:n())),
##    acct_group = n()
##  ) %>%
##  ungroup()
#join first account in HCAD to summed account in fixtures
#HCAD_fix_test <- left_join(HCAD_fixtures,HCAD_test,by="account")
#think about how to add geometry for each person to be separate in representations.

#http://pdata.hcad.org/Desc/2015/code_desc_real.txt
#in bldg_data -                                                                                                                                                           
#AP0                 Apt:  Effcncy Unit                                                                                                                                                                  
#AP1                 Apt:  1-Bedroom Unit                                                                                                                                                                
#AP2                 Apt:  2-Bedroom Unit                                                                                                                                                                
#AP3                 Apt:  3-Bedroom Unit                                                                                                                                                                
#AP4                 Apt:  4-Bedroom Unit 
#RMB                 Room:  Bedroom                                                                                                                                                                      
#RMT                 Room:  Total
#UNT                 Number of Apartment Units 

#https://hcad.org/hcad-resources/hcad-appraisal-codes/hcad-building-type-codes/
#perhaps still need to add logic for: (some come from improv_typ.x, some from improv_typ.y - res vs. real)
#NOT run - just in case we want to drill down
#fourplex4 <- HCAD %>% filter(improv_typ.x == 1004) #looks like each account has it listed as 8 bedrooms, instead of as 4 apts.
#1002 = duplex, 1003 = triplex, with same problems.


#TRY ST_IS VALID and MAKE VALID and see how it works on the null geometries



```

```{r add distances to censusblocks and then to HCAD}
#make more subunits
census_blocks_sf <- st_as_sf(censusblocks)
census_sub_blocks <- st_subdivide(census_blocks_sf,max_vertices = 8)
census_sub_blocks <- st_collection_extract(census_sub_blocks,type = c("POLYGON", "POINT", "LINESTRING"),warn = FALSE)
census_sub_blocks <- census_sub_blocks %>%
  group_by(geoid) %>%
  mutate(subgeoid = paste0(geoid,"_",100+row_number())) %>%
  ungroup()
census_sub_blocks$subgeoid_centroid <- st_centroid(census_sub_blocks$geometry)
#and add subgeoid to parcels
CensusSubBlockforHCADParcels <- st_within(HCAD$centroid, census_sub_blocks)
#unlist into vector
CensusSubBlockforHCADParcelsunlisted <- rapply(CensusSubBlockforHCADParcels,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
CensusSubBlockforHCADParcelsunlisted <- unlist(CensusSubBlockforHCADParcelsunlisted)
# add census geoid information to each parcel
HCAD_geom$subgeoid=census_sub_blocks$GEOID[CensusSubBlockforHCADParcelsunlisted]           
#and for 2019 - although should only need it for the crosswalks...
census_blocks19_sf <- st_as_sf(censusblocks_19)
census_sub_blocks19 <- st_subdivide(census_blocks19_sf,max_vertices = 8)
census_sub_blocks19 <- st_collection_extract(census_sub_blocks19,type = c("POLYGON", "POINT", "LINESTRING"),warn = FALSE)
census_sub_blocks19 <- census_sub_blocks19 %>%
  group_by(geoid_19) %>%
  mutate(subgeoid = paste0(geoid_19,"_",100+row_number())) %>%
  ungroup()
census_sub_blocks19$subgeoid_centroid <- st_centroid(census_sub_blocks19$geometry)
#assign each parcel to a subgeoid
CensusSubBlock19forHCADParcels <- st_within(HCAD$centroid, census_sub_blocks19)
#unlist into vector
CensusSubBlock19forHCADParcelsunlisted <- rapply(CensusSubBlock19forHCADParcels,function(x) ifelse(length(x)==0,9999999999999999999,x), how = "replace")
CensusSubBlock19forHCADParcelsunlisted <- unlist(CensusSubBlock19forHCADParcelsunlisted)
# add census geoid information to each parcel
HCAD$subgeoid=census_sub_blocks$GEOID[CensusSubBlock19forHCADParcelsunlisted]

#so we can have some reporting other direction, too, get population counts for the censusblocks.

HCAD_block_residence <- HCAD %>% 
  group_by(geoid) %>%
  summarize(residents_block = sum(num_residences,na.rm = TRUE)) 
  #need this to give numbers by sub, block, and tract.
HCAD_tract_residence <- HCAD %>% 
  group_by(tract) %>%
  summarize(residents_tract = sum(num_residences,na.rm = TRUE))
HCAD_subgeoid_residence <- HCAD %>% 
  group_by(subgeoid) %>%
  summarize(residents_subgeoid = sum(num_residences,na.rm = TRUE))

#for busstops #originally made in HCAD_geo, with ridership from April, 2022
busstops <- readRDS(paste0(houstondatadir,vintage,"/Metro_stops/busstops_blocks.RDS"))
busstops <- st_transform(busstops,crs=st_crs(HCAD_geom))
#distance to busstops for each block group
block_bus_distance <- st_distance(census_sub_blocks$subgeoid_centroid,busstops$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
bus_distance <- as.data.table(block_bus_distance)
census_sub_blocks$bus_distance <- apply(bus_distance[,1:length(colnames(bus_distance))], 1, min)

#distance to busstops for each block group
block_bus_distance19 <- st_distance(census_sub_blocks19$subgeoid_centroid,busstops$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
bus_distance19 <- as.data.table(block_bus_distance19)
census_sub_blocks19$bus_distance <- apply(bus_distance19[,1:length(colnames(bus_distance19))], 1, min)
censustracts_bus19a <- census_sub_blocks19 %>%
  group_by(TRACTCE) %>%
  mutate(tract_avg_min_bus = as.integer(mean(bus_distance))) %>%
  ungroup() %>%
  distinct(TRACTCE,tract_avg_min_bus)

```

```{r add census info back to right places in HCAD}

HCAD <- as.data.table(HCAD)
censusblocks <- as.data.table(censusblocks) #update its indices
censusblocks[,("geoid"):=GEOID]
HCAD <- as.data.table(HCAD)
censusblocks <- as.data.table(censusblocks)
#censusblocks[,("geoid"):=GEOID]
HCAD <- unique(HCAD,by=c("account","apt_id")) #check - this eliminated the expansion by apartment type - 
HCAD[geoid%in%unique(censusblocks$geoid),c("closest_food_distance","closest_food_account",
        "fast_food_distance_min","fast_food_account",
        "convenience_stores_distance","convenience_stores_account",
        "shop_center_distance","shop_center_account",
        "super_market_distance","super_market_account",
        "metro_stop_distance_min","metro_stop_BSID"):=
            censusblocks[.SD,c(list(food_source_distance_min),list(closest_food_account),
                               list(fast_food_distance_min),list(fast_food_account),
                               list(convenience_stores_distance),list(convenience_stores_account),
                               list(shop_center_distance),list(shop_center_account),
                               list(super_market_distance),list(super_market_account),
                               list(metro_stop_distance_min),list(metro_stop_BSID)),
                         on=("geoid")]]

HCAD_residences <- HCAD %>%
  filter(is.na(improv_typ_real)) #there are a significant number that are na on improv_typ, still, but chose to keep them for wholeness here
#cf: they seem to be things like commercial land in residential areas, some church properties, and other odds and ends
HCAD_businesses <- HCAD %>%
  filter(!is.na(improv_typ_real))
```

```{r make summaries for food insecurity and metro availability}
#have to make censustracts from HCAD_geo 
#remember that apartments are assumed to have 90% occupancy
censustracts[,("min_bus_distance_by_pop"):=HCAD[.SD,mean(),on=tract]]

#for busstops - more than 24 hours!!
#metro_stops <- readRDS(paste0(houstondatadir,"2022/metro_stops/busstops_tracts.RDS"))
#metro_stops <- st_transform(metro_stops,st_crs(HCAD))
#metro_stops_1k <- st_is_within_distance(HCAD$centroid,metro_stops$geometry,dist = 1000)

#metro_stops_distance <- st_distance(HCAD$centroid,metro_stops1$geometry,by_element = FALSE) #by element picks a vector and not a dense matrix - from closest to furthest - could choose a bunch of them
#metro_distance <- as.data.table(metro_stops_distance)
#HCAD$min_h_distance <- apply(h_distance[,1:90], 1, min)

#for crime - picked up 7/21/2022 - https://www.houstontx.gov/police/cs/crime-stats-archives.htm

saveRDS(HCAD,file = paste0(housingdir, vintage, "/HCAD_pts_",Sys.Date(),".RDS"))
saveRDS(HCAD_residences,file = paste0(housingdir, vintage, "/HCAD_residences_",Sys.Date(),".RDS"))
saveRDS(HCAD_businesses,file = paste0(housingdir, vintage, "/HCAD_businesses_",Sys.Date(),".RDS"))
```

